---
title: "Data Importing & Cleaning"
output: github_document
Name: Jasmin Martinez
Date: 03/18/2025
---

# Data Import
Data is stored in OneDrive, importing those data files into R to begin data processing and cleaning.
```{r include=FALSE}
library(haven)
time_to_diabetes = read_sas("C:/Users/jasmi/OneDrive/Desktop/Spring 2025/Thesis/THESIS_JRM2319/Mesa_TimeToDiabetes_20190604.sas7bdat")
MESA_E1 = read_sas("C:/Users/jasmi/OneDrive/Desktop/Spring 2025/Thesis/THESIS_JRM2319/MESAE1FinalLabel20240307.sas7bdat")
MESA_E5 = read_sas("C:/Users/jasmi/OneDrive/Desktop/Spring 2025/Thesis/THESIS_JRM2319/MESAe5_FinalLabel_20240213.sas7bdat")
```


# Cleaning time_to_diabetes datafile
- Observations: 6814  
- Variables kept-    
  - idno: Participant ID number  
  - dmage: Age at Diabetes diagnosis 
- Total number of missing values for **dmage**: 4917  
  - *What are we doing with missing values*: Missing values will be left as 'NA' because the regression models that will be ran into this project can handel the 'NA' automatically or I can explciitly state to omit those values. 
```{r}
library(dplyr)
time_to_diabetes = select(time_to_diabetes, idno, dmage)
sum(is.na(time_to_diabetes$dmage))
```


# Cleaning MESA_E1 datafile
```{r}
```

# Cleaning MESA_E5 datafile
```{r}

```

```{r}

```
