---
title: "Data Importing & Cleaning"
output: github_document
Name: Jasmin Martinez
Date: 03/18/2025
---

### Data Import
Data is stored in OneDrive, importing those data files into R to begin data processing and cleaning.

- Mesa_TimeToDiabetes_20190604.sas7bdat = **time_to_diabetes**  
- MESAE1FinalLabel20240307.sas7bdat = **MESA_E1**  
- MESAe5_FinalLabel_20240213.sas7bdat = **MESA_E5**  
```{r include=FALSE}
library(haven)
time_to_diabetes = read_sas("C:/Users/jasmi/OneDrive/Desktop/Spring 2025/Thesis/THESIS_JRM2319/Mesa_TimeToDiabetes_20190604.sas7bdat")
MESA_E1 = read_sas("C:/Users/jasmi/OneDrive/Desktop/Spring 2025/Thesis/THESIS_JRM2319/MESAE1FinalLabel20240307.sas7bdat")
MESA_E5 = read_sas("C:/Users/jasmi/OneDrive/Desktop/Spring 2025/Thesis/THESIS_JRM2319/MESAe5_FinalLabel_20240213.sas7bdat")
```


### Cleaning time_to_diabetes datafile
- **Time frame for data file**: Updated per follow-up or exam visit 
- **Starting # of Observations**: 6814 
- **Starting # of variables:** 12

- **Variables kept:** 2
<details>
  - idno: Participant ID number  
  - dmage: Age at Diabetes diagnosis
</details>
```{r include=FALSE}
library(dplyr)
time_to_diabetes = select(time_to_diabetes, idno, dmage)
```
- **Missing values**
```{r}
colSums(is.na(time_to_diabetes))
```
- *What am I doing with missing values*: Missing values will be left as 'NA' because the regression models that will be ran into this project can handle the 'NA' automatically or I can explicitly state to omit those values. However, this variable has a lot of missing data--we may or may not need this variable. Could be useful to show the average age of diabetes onset for the sample population, to be discussed.  

- **Final # of Observations:** 6814
- **Final # of variables:** 2

### Cleaning MESA_E5 datafile 
- **Time frame for data file**: April 2010-Feb 2012
- **Starting # of Observations**: 4716
- **Starting # of variables**: 297

- **Variables Kept:** 32
<details>
  - idno: Participant ID number 
  - ecomp5c: Exam 5 Completion Status
  - age1c: age at Exam 1 
  - agecat1c: age categories at Exam 1 
  - age5c: Age at Exam 5 
  - agecat5c: age categories at Exam 5
  - race1c: race/ethnicity
  - mexican1: Mexican, Chicano, Mexican-American
  - dominic1: Dominican 
  - puert1: Puerto Rican 
  - cuban1: Cuban
  - othhisp1: Other Hispanic 
  - lang1: Language spoken at Exam 1 
  - gender1: Gender 
  - site5c: Site at Exam 5
  - bmi5c: BMI 
  - cig5c: Cigarette smoking status
  - curalc5: Presently Drink Alcohol
  - dbp5c: Seated diastolic blood pressure 
  - sbp5c: seated systolic blood pressure
  - glucose5: Fasting glucose (mg/dl)
  - dm035c: Diabetes 2003 ADA Fasting Criteria 
  - ldl5: LDL cholesterol 
  - ldlcat5c: LDL cholesterol (categorical)
  - hdl5: HDL Cholesterol
  - hdlcat5c: HDL cholesterol (categorical)
  - chol5: Total cholesterol 
  - htn5c: Hypertension by JNC VI (1997) Creteria
  - income5: total gross family income, past 12 months 
  - exercm5c: Total intentional exercises- Min/Week
  - casisum5c: Total cognitive assessment score
  - numhhld5: # of people supported by family income
</details>
- **Missing Values:**
```{r include=FALSE}
library(dplyr)
MESA_E5 = select(MESA_E5, idno, ecomp5c, age1c, agecat1c, age5c, agecat5c, race1c, mexican1, dominic1, puert1, cuban1, othhisp1, lang1, gender1, site5c, bmi5c, cig5c, curalc5, dbp5c, sbp5c, glucose5, dm035c, ldl5, ldlcat5c, hdl5, hdlcat5c, chol5, htn5c, income5, exercm5c, casisum5c, numhhld5)
```
```{r}
colSums(is.na(MESA_E5))
```
Here, I am recoding dm035c (Diabetes 2003 ADA Fasting Criteria) from 4 categories into 3 categories. This recoding will combine "untreated diabetes" and "treated diabetes" into one category called "diabetes." 

Therefore, the three final categories of the new variables 'dm035c_recode' are 0: normal, 1: Impaired Fasting Glucose, and 2: Diabetes. 

Tabulations before recoding 
```{r}
table(MESA_E5$dm035c)
```

```{r}
library(dplyr)
MESA_E5 <- MESA_E5 %>%
  mutate(dm035c_recode = case_when(
    dm035c == 0 ~ 0,
    dm035c == 1 ~ 1,
    dm035c %in% c(2, 3) ~ 2
  ))
```
Tabulations after recoding 
```{r}
table(MESA_E5$dm035c_recode)

```

At this point, I'll only be removing participants who are not Hispanic/Latino (from race1c) and those who are missing values for the CASI assessment at Exam 5. Participants with missing values for exposures of interests will be potentially removed after merging Exam 1 and Exam 5. 

The reasoning for this is because variables across both Exam 1 and Exam 5 will be used in the creation of the acculturation score and the 'ever having diabetes' variables, therefore the full extent to missing-ness cannot be assessed until I get to that merged dataset.  


```{r echo=TRUE}
MESA_E5 = MESA_E5 %>% 
          filter(race1c == 4,
                 casisum5c != 'NA')
```
There are 999 Hispanic/Latino's in the sample, of these 15 have a missing CASI Score in Exam 5--we will exclude these participants. 

- **Final # of Observations:** 984
- **Final # of variables:** 32

```{r}
library(dplyr)
glimpse(MESA_E5)
```
## Cleaning MESA_E1 datafile
- **Time frame for data file**: July 2000-August 2002 
- **Starting # of Observations:** 6814
- **Starting # of variables:** 813

- **Variables kept:** 7
<details>
  - Idno: Participant ID number 
  - bth1: Birth Place-Self
  - ctrybth1: Birth country 
  - langhm1: Language Spoken at home 
  - lang1: Language Spoken at Exam 1 
  - educ1: education: highest level completed 
  - yrsalc1c: # of years drinking alcohol (current and former drinkers)
  - alcwk1c: # of drinks per week (current and former drinkers) 
</details>
```{r include=FALSE}
library(dplyr)
MESA_E1 = select(MESA_E1, idno, bth1, ctrybth1, langhm1, educ1, yrsalc1c, alcwk1c)

glimpse(MESA_E1)
```

I am keeping 15 variable. No participants will be removed from this data file because I'll be preforming a left join with the MESA_E5 data file being first and MESA_E1 being second, using idno. Therefore, the resulting dataset will only include the 984 Latinos who had CASI scores at Exam 5. 

- **Final # of observations:** 6814
- **Final # of variables:** 15

### Merging MESA_e5 (left) with MESA_e1 (right) to create *MESA_final*:
- **Final # of observations:** 984 
- **Final # of variables:** 38
```{r}
MESA_final = 
  left_join(MESA_E5, MESA_E1, by = "idno")
library(dplyr)
glimpse(MESA_final)
```

### Merging MESA_final(left) with time_to_diabetes (right) to create *MESA_final*:
- **Final # of observations:** 984 
- **Final # of variables:** 39
```{r}
MESA_final = 
  left_join(MESA_final, time_to_diabetes, by = "idno")
library(dplyr)
glimpse(MESA_final)

saveRDS(MESA_final, "MESA_final.rds")
```

### Assessing missing-ness in final sample:
```{r}
colSums(is.na(MESA_final))
```

