---
title: "Table 1"
output: github_document
Name: Jasmin Martinez
Date: 03/18/2025
---
```{r}
MESA_final <- readRDS("MESA_final.rds")
```

### Tabulation
#### Gender 
0: female  
1: Male
```{r}
table(MESA_final$gender1)
prop.table(table(MESA_final$gender1))
```
#### Nativity
1: One of 50 US states   
2: Puerto Rico   
3: Another country  
```{r}
table(MESA_final$bth1)
prop.table(table(MESA_final$bth1))
```
#### Nationality
Mexican, Chicano, Mexican-American   
0: No  
1: Yes  
```{r}
table(MESA_final$mexican1)
prop.table(table(MESA_final$mexican1))
```
Dominican   
0: N  
1: Y  
```{r}
table(MESA_final$dominic1)
prop.table(table(MESA_final$dominic1))
```
Puerto Rican   
0: No  
1: yes  
```{r}
table(MESA_final$puert1)
prop.table(table(MESA_final$puert1))
```
Cuban   
0: No  
1: Yes  
```{r}
table(MESA_final$cuban1)
prop.table(table(MESA_final$cuban1))
```
Other Hispanic   
0: No  
1: Yes   
```{r}
table(MESA_final$othhisp1)
prop.table(table(MESA_final$othhisp1))
```
#### Language spoken at home
1: ENGLISH  
2: SPANISH  
3: CANTONESE  
4: MANDARIN  
5: OTHER 
```{r}
table(MESA_final$langhm1)
prop.table(table(MESA_final$langhm1))
```

#### Language spoken at Exam 1 
1: English   
2: Spanish   
3: Chinese  
```{r}
table(MESA_final$lang1)
prop.table(table(MESA_final$lang1))
```

#### Geographic Location  
3: WFU = Winston-Salem, NC
4: COL = New York, NY  
5: JHU = Baltimore, MD  
6: UMN = Minneapolis, MN  
7: NWU = Chicago, IL  
8: UCLA = Los Angeles, CA  
```{r}
table(MESA_final$site5c)
prop.table(table(MESA_final$site5c))
```

#### Cigarette Smoking Status  
0: Never  
1: Former  
2: Current  
```{r}
table(MESA_final$cig5c)
prop.table(table(MESA_final$cig5c))
```

#### Hypertension  
0: No  
1: Yes  
```{r}
table(MESA_final$htn5c)
prop.table(table(MESA_final$htn5c))
```
#### Current Alcohol Consumption
```{r}
table(MESA_final$curalc5)
prop.table(table(MESA_final$curalc5))
```


#### Diabetes at Exam 5 
"Exam 5 Diabetes Mellitus by 2003 ADA Fasting Criteria Algorithm"   
0: Normal  
1: IFG ("impaired fasting glucose")  
2: Diabetes
```{r}
table(MESA_final$dm035c_recode)
prop.table(table(MESA_final$dm035c_recode))
```
_Question_: Should I create new variable combining Normal and IFG as "no diabetes" and untrested diabetes and treated diabetes as "diabetes"? 

### Mean & SDs  
Age  
```{r}
mean(MESA_final$age5c, na.rm = TRUE)
sd(MESA_final$age5c, na.rm = TRUE)
```
BMI  
```{r}
mean(MESA_final$bmi5c, na.rm = TRUE)
sd(MESA_final$bmi5c, na.rm = TRUE)
```
Average intentional exercises, minutes per week  
```{r}
mean(MESA_final$exercm5c, na.rm = TRUE)
sd(MESA_final$exercm5c, na.rm = TRUE)
```
- The mean and SD for Average intentional exercises, minutes per week seems a little off. Here is the distribution for the variable ot understand if there are any outliers or issues with the data:
```{r}
hist(MESA_final$exercm5c, xlab = "exercm5c")

boxplot(MESA_final$exercm5c, ylab = "exercm5c")

```
```{r}
Q1 = quantile(MESA_final$exercm5c, 0.25, na.rm = TRUE)
Q3 = quantile(MESA_final$exercm5c, 0.75, na.rm = TRUE)
IQR_value = IQR(MESA_final$exercm5c, na.rm = TRUE)

lower_threshold = Q1 - 1.5 * IQR_value
upper_threshold = Q3 + 1.5 * IQR_value

outliers = MESA_final$exercm5c[MESA_final$exercm5c < lower_threshold | MESA_final$exercm5c > upper_threshold]

print(outliers)

```


Average age of diabetes diagnosis, yrs
```{r}
mean(MESA_final$dmage, na.rm = TRUE)
sd(MESA_final$dmage, na.rm = TRUE)
```
Average LDL cholesterol 
```{r}
mean(MESA_final$ldl5, na.rm = TRUE)
sd(MESA_final$ldl5, na.rm = TRUE)
```
Average HDL cholesterol
```{r}
mean(MESA_final$hdl5, na.rm = TRUE)
sd(MESA_final$hdl5, na.rm = TRUE)
```
Average cognitive assessment score-CASI  
```{r}
mean(MESA_final$casisum5c, na.rm = TRUE)
sd(MESA_final$casisum5c, na.rm = TRUE)
```
Distribution of CASI and identifying the 10% 
```{r}
hist(MESA_final$casisum5c, xlab = "casisum5c")

CASI_cutoff = quantile(MESA_final$casisum5c, 0.10, na.rm = TRUE)
print(CASI_cutoff)
count_lower_10 <- sum(MESA_final$casisum5c <= CASI_cutoff, na.rm = TRUE)
print(count_lower_10)

count_lt74 = sum(MESA_final$casisum5c <= 74, na.rm = TRUE)
print(count_lt74)


```




